Rates of true polar wander in convecting planets

### 2015-08-28 tpw_rate.tex
Text

We are in a position to address the rates of true polar wander for a given convective moment $\bf E$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1}{3G}{\bf \Omega} \times {\bf \dot{\Omega} } = {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
\ifdetail

@@

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the quadrupole structure of the temperature field (see Appendix \ref{appendix:moments}).
Therefore, an estimate of one allows an estimate of the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
\label{eq:T_series}
\end{equation}
where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.

The integral in Equation \eqref{eq:temperature_fluctuations} is picking out a degree-two spherical harmonic in the lateral dimensions, and a degree-two polynomial in radius.
We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
\begin{equation}
\Lambda_{12} \sim \Gamma T_{202}
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

\begin{equation}
T_{\mathrm{degree 2}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
\end{equation}

@@

\begin{equation}
T_{\mathrm{degree 2}}(t) \sim \mathrm{Ra}^{-1}
\label{eq:degree_two_of_ra}
\end{equation}

For two-dimensional simulations shown below there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree2}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
This is illustrated in Figure \ref{fig:perturb}

@@

In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}.
We suggest, then, that a more vigorously convecting planet is much more likely to experience rapid TPW events.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
\caption{Top: Time series of principle moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.  Bottom: Time series of spin axis and mismatch angle $\theta$.  When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger.}
\label{fig:misfit}
\end{figure*}

@@

We may explicitly draw a connection between the moment of inertia of a rotating object and it's degree-two density structure.  The moment of inertia tensor may be written in index notation
\begin{equation}
I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) d^3 {\bf r}
\label{eq:inertia}
\end{equation}

where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  The degree-two density structure may be expressed as a quadrupole-moment tensor (cf. \citet{jackson1998classical}):

\begin{equation}
Q_{ij} = \int_V \rho \left( 3 r_i r_j - r_q r_q \delta_{ij} \right) d^3 {\bf r}
\end{equation}

The commutator of these tensors, $\bf QI - IQ$, is zero, indicating that they may be simultaneously diagonalized.

\ifdetail
We may demonstrate that the commutator is zero by using the definition of the commutator and expanding:
\begin{equation} 
\begin{aligned}
{\bf IQ-QI} = &I_{ik} Q_{kj} - Q_{ik} I_{kj} \\
 = &\int_V \rho \left( r_q r_q \delta_{ik} - r_i r_k \right) d^3 {\bf r} \int_V \rho \left( 3 r_k r_j - r_q r_q \delta_{kj} \right) d^3 {\bf r} \\
 &- \int_V \rho \left( r_q r_q \delta_{kj} - r_k r_j \right) d^3 {\bf r} \int_V \rho \left( 3 r_i r_k - r_q r_q \delta_{ik} \right) d^3 {\bf r} \\
 = &3 \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
 &-\delta_{kj} \delta_{ik} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 &-3 \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_k r_j d^3 {\bf r} \\
 &+\delta_{kj} \int_V \rho r_i r_k d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 - &3 \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
 &+\delta_{ik} \delta_{kj} \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 &+3 \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_i r_k d^3 {\bf r} \\
 &-\delta_{ik} \int_V \rho r_k r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 = \;&3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
 &+ \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 - &3 \int_V \rho r_q r_q d^3 {\bf r} \int_V \rho r_i r_j d^3 {\bf r} \\
 &- \int_V \rho r_i r_j d^3 {\bf r} \int_V \rho r_q r_q d^3 {\bf r} \\
 & = 0
\end{aligned}
\end{equation}
\else
\fi

We may therefore go into the principle coordinate system of both tensors to find
\begin{equation}
{\bf I} = {\bf 1} \begin{bmatrix}
\int_V \rho (y^2+z^z)\\
\int_V \rho (x^2+z^2) \\
\int_V \rho (x^2+y^2) 
\end{bmatrix}
\end{equation}

\begin{equation}
{\bf Q} = {\bf 1} \begin{bmatrix}
\int_V \rho (2 x^2 - y^2 - z^z)\\
\int_V \rho (2 y^2 - x^2 -z^2)\\
\int_V \rho (2 z^2 - x^2 - y^2) 
\end{bmatrix}
\end{equation}

where $\bf 1$ is the identity matrix.  If we let the diagonal elements of $\bf I$ be $\lambda_1, \lambda_2, \lambda_3$, we may rewrite $\bf Q$ as

\begin{equation}
{\bf Q} = {\bf 1} \begin{bmatrix}
(\lambda_2-\lambda_1) + (\lambda_3-\lambda_1)\\
(\lambda_1-\lambda_2) + (\lambda_3-\lambda_2)\\
(\lambda_1-\lambda_3) + (\lambda_2-\lambda_3)\\
\end{bmatrix}
\label{eq:maccullagh}
\end{equation}

which shows the connection between the degree-two component of the density field and the principle moments of inertia.  This may be seen as a form of MacCullagh's formula \citep{munk1960rotation}.

\section {TODO}

\begin{enumerate}
\item Make the argument about degree-two scaling much sharper
\item Discuss high-viscosity lower mantle
\item Earth-like parameters
\end{enumerate}

>>>>

We are in a position to address the rates of true polar wander for a given convective moment $\bf E$.
A considerable simplification occurs if we neglect secular changes in the rotation rate, and just consider changes in direction of the pole ($d \Omega^2 / dt = 2 {\Omega_i} \dot{ \Omega}_i = 0$)
Substituting Equation \eqref{eq:rotational_deformation} into Equation \eqref{eq:disequilibrium} we find
\begin{equation}
\frac{k_f a^5 T_1}{3G}{\bf \Omega} \times {\bf \dot{\Omega} } = {\bf \Omega} \times \left( {\bf E \cdot \Omega} \right)
\label{eq:liouville}
\end{equation}
Introducing a unit vector $\mitbf{\omega} = {\bf \Omega}/\|{\bf \Omega} \|$ and using Equation \eqref{eq:fluid_love} for $k_f$,  we may solve this equation for ${\bf \dot{\mitbf{\omega}} }$:
\begin{equation}
 \dot{\mitbf{\omega}}  = \frac{1}{(C-A)T_1} \left[ {\bf E \cdot \mitbf{\omega}} - \left( {\bf \mitbf{\omega} \cdot E \cdot \mitbf{\omega} } \right) \mitbf{\omega} \right]
\label{eq:liouville_ode}
\end{equation}
Note that the quantity in brackets is similar in form to the shear stress on a plane in classical elastostatics.
If we enter the coordinate system of the convective moment of inertia $\bf E$ with principal moments $\lambda_1 \le \lambda_2 \le \lambda_3$ and define the orientation of $\mitbf{\omega}$ with colatitude $\theta$ and longitude $\phi$ we can write this in a more illuminating form (after some tedious algebra):
\ifdetail

@@

\begin{equation}
 \rho_0{\bf \Omega \times \Omega \times r}= - \nabla P_0 + \rho_0 {\bf g}
\end{equation}

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-twoo part of the temperature field (see Appendix \ref{appendix:moments}).
Therefore, an estimate of one allows an estimate of the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi)
\label{eq:T_series}
\end{equation}
where the $T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$.

The integral in Equation \eqref{eq:temperature_fluctuations} is picking out degree-two spherical harmonics in the lateral dimensions, and low order polynomials in radius.
We therefore want to estimate the power in a single mode of the temperature expansion, allowing us identify
\begin{equation}
\Lambda_{12} \sim \Gamma T_{\mathrm{degree \textendash two}}
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

\begin{equation}
T_{\mathrm{degree \textendash two}}(t) \sim \frac{1}{N_{\mathrm{modes}}} \sim \left( \frac{d}{R}\right)^3
\end{equation}

@@

\begin{equation}
T_{\mathrm{degree \textendash two}}(t) \sim \mathrm{Ra}^{-1}
\label{eq:degree_two_of_ra}
\end{equation}

We performed a series of numerical simulations of mantle convection at different Rayleigh numbers to test this scaling.
We used the mantle convection software \texttt{ASPECT} \citep{kronbichler2012high}, based on the finite element library \texttt{deal.II} \citep{dealII81}. 
In order to test a wide range of Rayleigh numbers, we ran the simulations in a 2D annulus, tracking the eigenvalues of the 
moment of inertia tensor and integrating Equation~\eqref{eq:liouville_ode} in time.
For the 2D simulations there is a reduced dimensionality when calculating the number of modes,
so $N_\mathrm{modes} \sim \left(R/d \right)^2$.  This leads us to a scaling of $T_{\mathrm{degree\textendash two}} \sim \mathrm{Ra}^{-2/3}$, which is used in the simulations shown in Figure~\ref{fig:eigengap}.
This result has a simple interpretation.
As the Rayleigh number of the system increases, the smallest lengthscale of convective features gets smaller.
The total power in the temperature field is spread across a larger spectrum, leaving less total power for the degree-two part, which is what drives TPW.

@@

That is to say, if there is a large difference between the eigenvalues, this stabilizes axes to perturbations.  
If, however, there is a small difference between the eigenvalues (i.e., they are nearly degenerate), then perturbations can cause large rotations of the principle axes.
This is illustrated schematically in Figure \ref{fig:perturb}

@@

In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
and if it does go to zero, the wander does not need to be $90^\circ$.

Figure~\ref{fig:misfit} shows a representative timeseries for annular convection, where 
we track the eigenvalues of the moment of inertia and integrate Equation~\eqref{eq:liouville_ode} in time.
Since it is a 2D model, the spin axis may be represented by a single angle.
When the eigengap gets small, the misfit angle becomes much larger, and the rate of polar wander becomes much faster.
At $\sim$1 Gyr the eigengap goes to zero, and the misfit angle goes to approximately $90^\circ$, 
an IITPW event. But there are several other events where the eigengap becomes small, and there are 
still large TPW events associated with them.

We suggest, then, that $90^\circ$ IITPW events are simply a special case of a broad class of events which can occur
when the principal moments are close to each other.  Furthermore, a more vigorously convecting planet is much more likely to experience rapid TPW events
as it has a lower characteristic gap between principal moments and the gap is much more likely to go to zero.

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{figures/misfit.pdf}
\caption{Top: Time series of principle moments for 2D annular convection at $\mathrm{Ra}\sim10^8$.  Bottom: Time series of spin axis and mismatch angle $\theta$.  When the two moments are close to each other (small eigengap), the mismatch angle becomes large, and the rate of polar wander is significantly larger. At $\sim$1 Gyr the gap goes to zero and there is a nearly $90^\circ$ TPW event. However, there are several other large TPW events which happen when the eigengap is small.}
\label{fig:misfit}
\end{figure*}

@@

We may explicitly draw a connection between the moment of inertia of a rotating object and it's degree-two density structure.  The moment of inertia tensor may be written in index notation
\begin{equation}
I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
\label{eq:inertia}
\end{equation}
where $\bf r$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
It is useful to enter the principal axes of the moment of inertia:
\begin{equation}
{\bf I} = \bf{1} \begin{bmatrix}
\lambda_1 \\ \lambda_2 \\ \lambda_3
\end{bmatrix} = 
{\bf 1} \begin{bmatrix}
\int_V \rho (y^2+z^z) dV\\
\int_V \rho (x^2+z^2) dV\\
\int_V \rho (x^2+y^2) dV
\end{bmatrix}
\end{equation}
where $\bf 1$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
From equation~\eqref{eq:milankovitch} we see that the important quantities are the differences between the 
principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
The relevant (fully normalized) harmonics are, in Cartesian coordinates:
\begin{equation}
\begin{aligned}
Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}
\end{aligned}
\end{equation}
Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
\ifdetail
It is useful to define unnormalized verions of the spherical harmonics as intermediate products:
\begin{equation}
\begin{aligned}
Y_{20}^* &= 2 z^2 - x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{5} } r^2 Y_{20} \\ 
Y_{22}^* &= x^2 - y^2 &= 4 \sqrt{ \frac{\pi}{15} } r^2 Y_{22}
\end{aligned}
\end{equation}
These make it easy to rewrite the differences between the $\lambda_n$:
\begin{equation}
\begin{aligned}
(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
\end{aligned}
\end{equation}
We may split this into integrations in radius and integrations over the surface of a sphere
\begin{equation}
\begin{aligned}
(\lambda_2 - \lambda_1) &= \int_V \rho Y_{22}^* \, dV \\
(\lambda_3 - \lambda_1) &= \frac{1}{2} \int_V  \rho \left( Y_{22}^* - Y_{20}^* \right) dV \\
(\lambda_3 - \lambda_2) &= -\frac{1}{2} \int_V \rho \left( Y_{22}^* + Y_{20}^*  \right) dV \\
\end{aligned}
\end{equation}
Substituting back in for the normalized spherical harmonics we find
\fi
\begin{equation}
\begin{aligned}
(\lambda_2 - \lambda_1) &= 4 \sqrt{\frac{\pi}{15} } \int_V   \rho r^2  Y_{22} \,dV \\
(\lambda_3 - \lambda_1) &= 2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} - \sqrt{3} Y_{20} \right) dV \\
(\lambda_3 - \lambda_2) &= -2 \sqrt{ \frac{\pi}{15} } \int_V  \rho r^2 \left( Y_{22} + \sqrt{3} Y_{20} \right) dV \\
\end{aligned}
\end{equation}
Up to the normalization constants, these expressions are idential to multipole expansions, 
picking out the degree-two part of the density field laterally, and low-order polynomials radially.
When density is a function of temperature, then we may directly identify this with the same expansion of the temperature field.
